---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data is in the form of a csv file inside a zip folder. So, the following steps have been followed to load the data.
1. Unzip the file
2. Read the ensuing csv file using read.csv

```{r chunk1, echo = TRUE}
unzip("activity.zip")
activityData <- read.csv("activity.csv", header = TRUE)
print(head(activityData))
```

## What is mean total number of steps taken per day?
1. The total number of steps per day can be found by running *tapply()* on *steps* over the margin *date*. To ensure that missing values are not considered, na.rm has been set to TRUE

```{r chunk2, echo = TRUE}
stepsPerDay <- tapply(activityData$steps, activityData$date, sum, na.rm = TRUE)
```

2. The histogram, basically, is a frequency chart of the number of steps per day. To plot the histogram, first the number of levels of dates (which is a factor), are found.This is just to get the number of bins needed in the histogram.Then, the histogram of *stepsPerDay* is plotted.

```{r chunk3, echo = TRUE}
lenDates <- length(levels(activityData$date))
hist(stepsPerDay, lenDates, col = "deepskyblue")
```

3. The mean value of steps per day, is just the mean of the *stepsPerDay* variable. The median is also found below.

```{r chunk4, echo = TRUE}
meanStepsPerDay <- mean(stepsPerDay)
printMean <- paste("Mean Steps per day = ", meanStepsPerDay)
cat(printMean)
medianStepsPerDay <- median(stepsPerDay)
printMedian <- paste("Median of Steps Per Day is ", medianStepsPerDay)
cat(printMedian)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
